PAMELA: an annotation-based Java Modeling Framework

### 2021-02-17 ResponseLetter.tex

\begin{response}{gives "git clone git@github.com:openflexo-team/pamela.git", which (even when using my own username) does not work. However, "git clone https://github.com/openflexo-team/pamela.git" immediately does. Why not state the latter?
}

We apologize for the inconvenience. We have replaced "git clone git@github.com:openflexo-team/pamela.git" for "git clone https://github.com/openflexo-team/pamela.git" as suggested.
\end{response}

\begin{response}{how to actually build the environment is not stated for this option.}
We apologize for the inconvenience. The PAMELA documentation and web site have been updated in order to solve the problem.
\end{response}

\begin{response}{Run tests Unit tests located in src/test/java in pamela-core are a good starting point to discover the framework." No further information is given. Apparently, the user has to figure out him/herself how to run the tests. Documentation is lacking here.
}
We agree with the reviewer. The PAMELA documentation and web site have been updated in order to improve tests understanding (see section 2.2 of "Getting started"). 
\end{response}

@@

\begin{response}{how to use the tool is not describe properly. It is not clear if the developers need to write all the annotations. If an application is built with PAMELA, what is the link between the interfaces and the concrete classes. Indeed, at one point, code corresponding to the behavior will be needed; and not only the accessors to navigate the structure of the model.}
We have modified Section 2 to clarify that standard PAMELA users will reuse existing annotations which are accompanied with default implementations in the PAMELA interpreter. Additional details are given in Sections 3 and 4 (and in the online documentation: \url{https://pamela.openflexo.org/SNAPSHOT/pamela-core/3-metaprogramming_support.html}) See Listing 3 for an example of how the default implementation can be replaced by custom ones.
\end{response}

\begin{response}{how does such "code" evolve with time? Authors have used PAMELA for years, how such applications resist with time?}
PAMELA helps factorizing code corresponding to recurrent concepts. In that sense its use does reduce the size and duplication of code, decreases the number of bugs and improves the code structure. All this makes final applications that use PAMELA (more) easily maintainable.
We briefly discuss in Section 5.1. how integrating PAMELA in the OpenFlexo tool helped removing duplicated code and eliminating bugs in the latter.
\end{response}

\begin{response}{section 2 is too abstract to be very useful}
Please, see the two first answers above where we explain how we have extended Section 2.
\end{response}

\begin{response}{p6 "according to PAMELA semantics" but it is never explained.}
We have removed this reference to PAMELA semantics in the paper. Our intention was to refer to the PAMELA metamodel.
\end{response}

### 2021-03-22 architecture.tex
Added Todo-Review2.tex to keep track of changes

PAMELA is composed of one design time component (the PAMELA metamodel, plus a number of predefined annotations) presented in the next section and one runtime component (the PAMELA interpreter), described in the section \ref{sub:RunTime}.

>>>>

PAMELA is composed of one design time component (the PAMELA metamodel, plus a number of predefined annotations) presented in the next section and one runtime component (the PAMELA interpreter), described in Section \ref{sub:RunTime}.

### 2021-03-22 pamelaProcess.tex
Added Todo-Review2.tex to keep track of changes

We advocate for a strong coupling between model and sources code, to give architects and developers a way to both interact during the whole development cycle. PAMELA is an annotation-based Java modeling framework providing a smooth integration between model and code, without code generation nor externalized model serialization. Instead of generating the code, the API (mostly Java interfaces with abstract method declarations) is locally executed (interpreted). The idea is to avoid separation between modeling and code to facilitate consistency management and to /avoid round-trip issues. Figure \ref{fig:PamelaVision} summarizes the PAMELA architecture and is more detailed thereafter.

@@

Coupling model and code into the same artifact opens new ways of programming. The classical (metadata enabled) programming process relies on \emph{programmers} that produce code reusing pre-existing modeling concepts. These concepts are implemented by \emph{modelers} that provide the right annotations the programmers use. This is, for instance, the process followed by Jakarta EE (JEE) developers reusing JEE specific annotations. The evolution rhythm between models and code is low. This programming way is still possible with PAMELA, but we allow the ability to reach a high evolution rhythm when the programmer also becomes the modeler. In fact, when the programmer identifies a pattern, an abstraction, a generalization, s/he can use PAMELA to develop and capitalize on this abstraction by increasing PAMELA metamodel.

@@

The developed metamodels are implemented by annotations that rely on Java/JVM entities and mechanisms. They include consistency checking, which constrain their use and helps the programmer to avoid inconsistencies or errors. We have first experimented their use with setter/getter to define POJO entities, with traits to implement multiple inheritance or with roles and rules to set security rules on classes.

>>>>

We advocate for a strong coupling between model and sources code, to give architects and developers a way to both interact during the whole development cycle. PAMELA is an annotation-based Java modeling framework providing a smooth integration between model and code, without code generation nor externalized model serialization. Instead of generating the code, the API (mostly Java interfaces with abstract method declarations) is locally executed (interpreted). The idea is to avoid separation between modeling and code to facilitate consistency management and to avoid round-trip issues. Figure \ref{fig:PamelaVision} summarizes the PAMELA architecture and is more detailed thereafter.

@@

Coupling model and code into the same artifact opens new ways of programming. The classical (metadata enabled) programming process relies on \emph{programmers} that produce code reusing pre-existing modeling concepts. These concepts are implemented by \emph{modelers} that provide the right annotations the programmers use. This is, for instance, the process followed by Jakarta EE (JEE) developers reusing JEE specific annotations. The evolution rhythm between models and code is low. This programming way is still possible with PAMELA, but we allow the ability to reach a high evolution rhythm when the programmer also becomes the modeler. In fact, when the programmer identifies a pattern, an abstraction, a generalization, s/he can use PAMELA to develop and capitalize on this abstraction by extending PAMELA's metamodel.

@@

The developed metamodels are implemented by annotations that rely on Java/JVM entities and mechanisms. They include consistency checking, which constrain their use and helps the programmer to avoid inconsistencies or errors. We have first experimented their use with setters/getters to define Plain Old Java Objects (POJO), with traits to implement multiple inheritance or with roles and rules to set security rules on classes.
